Synthesis of Linear Functional Programs

### 2021-07-20 article/article.tex
ADT poly + memoization tweaks

\mypara{Polymorphic ADTs}

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
especially noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications:
%TODO: ignore constraints if type isnnt existential
%
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
        \hline
        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
        \hline
    \end{tabular}
\end{center}
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality for polymorphic types.

\mypara{Debugging} The rules described in the formal
system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
retrospective, however, when faced with a program that doesn't terminate, what
went wrong is not so clear. While we were able to figure out a solution for the
first couple of "infinite examples", the following problems were much harder,
taking sometimes full days to solve (and as of now, some still remain unsolved,
e.g. List (List a) $\lolli$ List a). To make the debugging experience better, we
developed a tracing system that prints information whenever a rule is applied,
alongside the stack of rules applied up until that one. This improved our
debugging speed considerably, especially when put together with a correct
derivation done by hand, with which we could match what was happening against
what was supposed to happen.

Unfortunately, it seems that sometimes in the process of backtracking, stack
information is "lost", and the trace logs lose useability.

>>>>

\mypara{Polymorphic ADTs} To allow type parameters and the use of universally quantified type
variables in ADT constructors, we must guarantee that the
\textsc{adt-init} rule can unify the type parameters and that when constructing or
destructing an ADT, type variables in constructor parameters are substituted by
the actual type (i.e. to construct \mintinline{haskell}{List Int} with
\mintinline{haskell}{data List a = Cons (a *
List a)}, we wouldn't try to synthetise \mintinline{haskell}{(a * List a)},
but rather \mintinline{haskell}{(Int * List Int)}). To unify $T_{\overline\alpha}$ with $T_{\overline\beta}$,
the sets of type parameters $\overline\alpha$ and $\overline\beta$ must satisfy $|\overline\alpha| = |\overline\beta|$
together with $\forall i\ 0 \leq i \land i < |\overline\alpha| \land
\textsc{unify}(\overline\alpha_i \mapsto \overline\beta_i)$. The constructor
type substitution needn't be explicit in the rule:
\begin{mathpar}
    \infer*[right=(adt-init)]
    {\textsc{unify}(T_{\overline\alpha} \mapsto T_{\overline\beta}, \Theta)}
    {\Theta/\Theta,T_{\overline\alpha} \mapsto T_{\overline\beta},\Rho; \Gamma; \Delta/\Delta'; x{:}T_{\overline\alpha} \Downarrow\ \vdash x :
    T_{\overline\beta}}
\end{mathpar}

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
quite noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications. For example, if we add a new function
\emph{read} to the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take around
$5s$, while synthesis without memoization doesn't terminate even after
$15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some examples.

\mypara{Debugging} The rules described in the formal
system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
retrospective, however, when faced with a program that doesn't terminate, what
went wrong is not so clear. While we were able to figure out a solution for the
first couple of "infinite examples", the following problems were much harder,
taking sometimes full days to solve. To make the debugging experience better, we
developed a tracing system that prints information whenever a rule is applied,
alongside the stack of rules applied up until that one. This improved our
debugging speed considerably, especially when put together with a correct
derivation done by hand, with which we could match what was happening against
what was supposed to happen.
